---
title: "Draft Work"
output: html_document
date: "2023-03-07"
---

```{r setup, include=FAL
```{r}
# ###Format data
blind_spot <- c(26, 35) # define blind spot
dat_model <- dat[order(dat$location), ] # sort by location
dat_model <- dat[!dat$location %in% blind_spot, ] # remove blind spot locations
Y <- dat_model$dls # define observed outcome data
Time <- unique(dat_model$years)
  W <- HFAII_Queen[-blind_spot, -blind_spot] # visual field adjacency matrix
DM <- GarwayHeath[-blind_spot] # Garway-Heath angles
TimeDist <- abs(outer(Time, Time, "-"))
TimeDistVec <- TimeDist[lower.tri(TimeDist)]
minDiff <- min(TimeDistVec)
maxDiff <- max(TimeDistVec)
PhiUpper <- -log(0.01) / minDiff # shortest diff goes down to 1%
PhiLower <- -log(0.95) / maxDiff # longest diff goes up to 95%
Hypers <- list(Delta = list(MuDelta = c(3, 0, 0), OmegaDelta = diag(c(1000, 1000, 1))),
               T = list(Xi = 4, Psi = diag(3)),
               Phi = list(APhi = PhiLower, BPhi = PhiUpper))
Starting <- list(Delta = c(3, 0, 0), T = diag(3), Phi = 0.5)
Nu <- length(Time) # calculate number of visits
Tuning <- list(Theta2 = rep(1, Nu), Theta3 = rep(1, Nu), Phi = 1)
MCMC <- list(NBurn = 10000, NSims = 250000, NThin = 25, NPilot = 20)
PlotAdjacency(W = W, DM = DM, zlim = c(0, 180), Visit = NA,
              main = "Garway-Heath dissimilarity metric\n across the visual field")

reg.STBDwDM <- STBDwDM(Y = Y, DM = DM, W = W, Time = Time,
                       Starting = Starting,
                       Hypers = Hypers,
                       Tuning = Tuning,
                       MCMC = MCMC,
                       Family = "tobit",
                       TemporalStructure = "exponential",
                       Distance = "circumference",
                       Weights = "continuous",
                       Rho = 0.99,
                       ScaleY = 10,
                       ScaleDM = 100,
                       Seed = 54)

#encoded as negative
```
```{r}
dat = dat %>% 
  mutate(dls_scaled = dls/10)

dat$dls_scaled[dat$dls_scaled == -0.1] <- 0
Time_inc <- (unique(dat$age) / 365)
datmodel_o <- data.frame(Y = (dat$dls) / 10) # create data frame with scaled data
M <- dim(W)[1] # number of locations
TimeDist <- as.matrix(dist(Time))
BPsi <- log(0.025) / -min(TimeDist[TimeDist > 0])
APsi <- log(0.975) / -max(TimeDist)
K <- 10
O <- 1
Hypers <- list(Sigma2 = list(A = 0.001, B = 0.001),
               Kappa = list(SmallUpsilon = O + 1, 
                            BigTheta = diag(O)),
               Delta = list(A1 = 1, A2 = 20),
               Psi = list(APsi = APsi, BPsi = BPsi),
               Upsilon = list(Zeta = K + 1, Omega = diag(K)))
Starting <- list(Sigma2 = 1,
                 Kappa = diag(O),
                 Delta = 2 * (1:K),
                 Psi = (APsi + BPsi) / 2,
                 Upsilon = diag(K))
Tuning <- list(Psi = 1)
# MCMC <- list(NBurn = 1000, NSims = 1000, NThin = 2, NPilot = 5)
# reg.bfa_sp <- bfa_sp(dls_scaled ~ iop, data = dat, dist = W, time = sort(Time_inc),  K = 10, 
#                      starting = Starting,
#                      hypers = Hypers,
#                      tuning = Tuning,
#                      mcmc = MCMC,
#                      L = Inf,
#                      family = "tobit",
#                      trials = NULL,
#                      temporal.structure = "exponential",
#                      spatial.structure = "discrete",
#                      seed = 54, 
#                      gamma.shrinkage = TRUE,
#                      include.space = TRUE,
#                      clustering = TRUE)


```

