---
title: "Rob.qmd"
format: html
editor: visual
---

## EDA

Taken from main doc

```{r}
packages <- c('tidyverse', 'spBFA','womblR', 'patchwork', 'tidymodels')
install.packages(setdiff(packages, rownames(installed.packages())))
library(patchwork)
library(tidymodels)
library(spBFA)
library(tidyverse)
```

```{r}
#| label: generate-data
source(file = "case3_convert_to_right_eye.R")
```

```{r}
Bryan2013 <- read_csv("~/Downloads/case3-team06/glaucoma_data/Bryan2013.csv")
Erler2014 <- read_csv("~/Downloads/case3-team06/glaucoma_data/Erler2014.csv")
Patients <- read_csv("~/Downloads/case3-team06/glaucoma_data/Patients.csv")
VFPoints <- read_csv("~/Downloads/case3-team06/glaucoma_data/VFPoints.csv")
VisualFields <- read_csv("~/Downloads/case3-team06/glaucoma_data/VisualFields.csv")

```

# Methodology

Describe methods in sufficient detail for replication.

## Exploratory data analysis

EDA is important!

```{r}
Patients %>%
  group_by(SEX) %>%
  summarize(n.sex=n())
```

```{r}
plot1 <- ggplot(dat, aes(x = log(dls))) +
  geom_histogram(binwidth = .01)
plot2 <- ggplot(dat, aes(x = iop)) +
  geom_histogram(binwidth = 1)
plot1 + plot2
```

```{r}
blind_spot <- c(26, 35) # define blind spot
W <- HFAII_Queen[-blind_spot, -blind_spot] # HFA-II visual field adjacency matrix
library(igraph)
adj.graph <- graph.adjacency(W, mode = "undirected") 
plot(adj.graph)
```

New EDA

```{r}
#| label: exploring age through EDA

vf_od |>
  ggplot(
    aes(x = AGE, 
        y = TOTAL_DEVIATION,
        color = SEX)
  ) + 
  geom_point(alpha = .01)

#males are blue, females red
```

Any Change between age looks just to be that we have older females in study. Otherwise there is a very clear delineation between a lot of deviation and not much at all with minimal density in the middle. No crazy correlation with age. Bi-modal density for deviation shown below.

```{r}
#| deviation density
VisualFields |>
  ggplot(
    aes(x = MD
  )) + 
  geom_density() + 
  labs(title = "Distribution of Mean Deviation for All Patients",
       x  = "Mean Deviation from 'Normal' Vision",
       y = "") + 
  theme_minimal() + 
  theme(axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank())
```


```{r}
library(womblR)

Patient4_eye_OD <- dat |>
  filter(id == 7)

PlotVfTimeSeries(Y = Patient4_eye_OD$dls,
                 Location = Patient4_eye_OD$location,
                 Time = Patient4_eye_OD$time,
                 main = "Visual field sensitivity time series \n at each location for right eye of patient 4",
                 xlab = "Days from baseline visit",
                 ylab = "Differential light sensitivity (dB)",
                 line.reg = FALSE)

```




```{r}
#| deviation by region


vf_od |>
  group_by(Location) |>
  summarise(mean_dev = mean(TOTAL_DEVIATION)) |>
    ggplot(
      aes(x = Location, 
        y = -mean_dev)) + 
    geom_col()
```

Looks to be a stairstep pattern which may be interesting to dig into further.

```{r}
#| deviation by region


vf_od |>
  group_by(Location) |>
  summarise(mean_dev = mean(TOTAL_DEVIATION)) |>
    ggplot(
      aes(x = Location, 
        y = -mean_dev)) + 
    geom_col()


```

```{r}
VisualFields_EDA <- dat |>
  mutate(early = if_else(id %in% c(3, 4, 12, 13, 27, 30, 34, 43, 46, 47, 61, 72, 76, 90, 98, 102, 114, 125, 126, 128), "Young", "Old"))
                                 
```

```{r}
VisualFields_EDA |>
  ggplot(
    aes(x = time,
        y = md)) + 
  geom_point(alpha = .01) + 
  geom_smooth(method = "lm") + 
  annotate("text",x=4,y=5,size = 3, label=(paste0("slope==",coef(lm(VisualFields_EDA$md~VisualFields_EDA$time))[2]))) +
  facet_wrap(~ early) + 
  theme_minimal()
```

```{r}
#| layout-ncol: 2

VisualFields_EDA_young <- VisualFields_EDA |>
  filter(early == "Young")


VisualFields_EDA_old <- VisualFields_EDA |>
  filter(early == "Old")


VisualFields_EDA_young |>
  ggplot(
    aes(x = time,
        y = md)) + 
  geom_point(alpha = .01, color = "lightgreen") + 
  geom_smooth(method = "lm") + 
  annotate("text",x=4,y=5,size = 3, label=(paste0("slope=",round(coef(lm(VisualFields_EDA_young$md~VisualFields_EDA_young$time))[2],3)))) + 
  labs(title = "Mean Deviation for Younger Patients Over Time",
     subtitle = "Patients Who Got Glaucoma Before Age 46",
     y = "Mean Visual Deviation\nAcross Points",
     x = "Years") + 
  theme_minimal()


VisualFields_EDA_old |>
  ggplot(
    aes(x = time,
        y = md)) + 
  geom_point(alpha = .001) + 
  geom_smooth(method = "lm") + 
  annotate("text",x=4,y=5,size = 3, label=(paste0("slope=",round(coef(lm(VisualFields_EDA_old$md~VisualFields_EDA_old$time))[2],3)))) + 
labs(title = "Mean Deviation for Older Patients Over Time",
     subtitle = "Patients Who Got Glaucoma After Age 45",
     y = "Mean Visual Deviation\nAcross Points",
     x = "Years") + 
theme_minimal()

```

